---
title: "BIEN Biostatistics HELP for boxplots and t-tests"
output: html_notebook
---

## Trait Data:  Using BIEN data to practice boxplots and t-tests

We learned to download trait data. First we downloaded data for the entire genus Artocarpus:

```{r}
# Create data frame of breadfruit traits
breadfruit_traits <- BIEN_trait_genus(genus = "Artocarpus")
str(breadfruit_traits)
head(breadfruit_traits)
```

And I suggested that you download data for just one trait, diameter at breast height (or DBH).

```{r}
Breadfruit_DBH <- BIEN_trait_traitbygenus(trait = "diameter at breast height (1.3 m)", genus = "Artocarpus")
#head(Breadfruit_DBH)
#tail(Breadfruit_DBH)
#str(Breadfruit_DBH)
names(Breadfruit_DBH)
#You got a lot of variables in addition to the DBH trait data, which has the column name "trait_value"
```

Then we trimmed the big downloaded dataframe to save just the first five columns. 

As data are initially downloaded from BIEN, they come down as character data, not numeric data. So, there is code below to do two things:

1. Re-name trait_value as DBH

2. Change the variable DBH from a character variable to a numeric variable.

In the trimmed file, the trait is still called "trait_value".

We extracted DBH from the trimmed file, and it now exists as just a stand-alone variable. DBH is not part of the dataframe Breadfruit_DBH_small.

```{r}
Breadfruit_DBH_small = Breadfruit_DBH[c(1:5)]
head(Breadfruit_DBH_small)
#tail(Breadfruit_DBH_small)
Breadfruit_DBH_small = as.data.frame(Breadfruit_DBH_small)
#We can see that our new dataframe is a dataframe
class(Breadfruit_DBH_small)
#But what type of variable is the trait value? It should be numeric, but it's not. It comes from the BIEN database as a character variable. 
class(Breadfruit_DBH_small$trait_value)
#So, in addition to renaming it
DBH = Breadfruit_DBH_small$trait_value
class(DBH)
#We also need to change it from a character variable to a numeric variable
DBH = as.numeric(Breadfruit_DBH_small$trait_value)
class(DBH)
```

Note two things are being done below:

1. Loading a library used for making boxplots. 

2. asking for a boxplot just with our stand-alone variable DBH


```{r}
#We learned to use the package lattice, for making boxplots and histograms of data
library(lattice)
boxplot(DBH~scrubbed_species_binomial, data=Breadfruit_DBH_small)
```

Then we downloaded another variable for the whole genus, whole plant height  	

```{r}
Breadfruit_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Artocarpus")
names(Breadfruit_WPH)
#Breadfruit_WPH
#Notice again that the trait is in there just as "trait_value"
```

One again, slimming down the dataframe is a good idea. 

And then recycle your code from above and create a new stand-alone variable called WPH, making sure it is converted into a numeric variable.
```{r}
Breadfruit_WPH_small = Breadfruit_WPH[c(1:5)]
head(Breadfruit_WPH_small)
Breadfruit_WPH_small = as.data.frame(Breadfruit_WPH_small)
class(Breadfruit_WPH_small)
head(Breadfruit_WPH_small)
class(Breadfruit_WPH_small$trait_value)
WPH = as.numeric(Breadfruit_WPH_small$trait_value)
class(WPH)
```


We can figure out the mean values of DBH and WPH, which is the mean for the entire genus. 

Since we defined DBH and WPH as stand-alone variables, we can use simplified code:

```{r}
mean(DBH)
mean(WPH)
```


Next, we switched over downloading for just one species at a time, starting with Artocarpus nitidus: 

note we are using a different function in the BIEN package: BIEN_trait_traitbyspecies

```{r}
#library(BIEN)
Anitidus_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Artocarpus nitidus")
head(Anitidus_DBH)
#tail(Anitidus_DBH)
```

and the data are similarly formatted, and so we can recycle similar code to slim down the dataframe: 

Again, we also created a new stand-alone variable for this species, called DBHn (for the species A. nitidus)

Always, data from BIEN must be converted to a numeric variable before doing plotting or biostatistics.

```{r}
Anitidus_DBH_small <-Anitidus_DBH[c(1:5)]
class(Anitidus_DBH_small)
head(Anitidus_DBH_small)
class(Anitidus_DBH_small$trait_value)
DBHn = Anitidus_DBH_small$trait_value
class(DBHn)
#We only named a new stand-alone variable, we didn't change it from character to numeric
DBHn = as.numeric(Anitidus_DBH_small$trait_value)
class(DBHn)
#Now it's changed to numeric
```


```{r}
hist(DBH)
boxplot(DBH)
#You get just one "box" because there is no grouping - this is all data for just one genus.
hist(DBHn)
boxplot(DBHn)
#Similarly, you get just one "box" because this is data for just one species.
```


Repeating everything for another species, Artocarpus hirsutus: 

```{r}
Ahirsutus_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Artocarpus hirsutus")
head(Ahirsutus_DBH)
#tail(Ahirsutus_DBH)
#class(Ahirsutus_DBH)
```


```{r}
Ahirsutus_DBH_small = Ahirsutus_DBH[c(1:5)]
class(Ahirsutus_DBH_small)
head(Ahirsutus_DBH_small)
DBHh = Ahirsutus_DBH_small$trait_value
class(DBHh)
DBHh = as.numeric(Ahirsutus_DBH_small$trait_value)
class(DBHh)
```

We could plot of DBHh just for the species Artocarpus hirsutus


```{r}
hist(DBHh)
boxplot(DBHh)
#Similarly, you get just one "box" because this is data for just one species.
```


## A simple way to do a t-test: compare DBHn with DBHh

Now we have created two variables, for two different species: :

1. DBHn is data for A. nitens

2. DBHh is data for A. hirsutus

You can make a boxplot in this simple way for these two variables. 

```{r}
boxplot(DBHh, DBHn)
```

You could do this for more than two variables, if you had defined them. Just for fun, let's compare the two species and the whole genus:

```{r}
boxplot(DBHh, DBHn, DBH)
```



Also, we can then run a t-test by simply assigning one of our stand-alone numeric variables as x and the other as y. T-test is always for just two groups, so we couldn't add more variables as we could for the boxplot. 

```{r}
#We just defined one variable in vector form, which is DBHn (all the DBH values for A. nitens)
DBHn
#We also defined a similar variable in vector form, DBHh (which is all the DBH variables for A. hirsutus)
DBHh
#we can run a t-test just by defining one variable as x, the other as y. 
ttest_result1 = t.test(x=DBHn, y=DBHh, paired=FALSE, alternative="two.sided")
ttest_result1
```

If we reverse the order, we get the same result:

```{r}
#We just defined one variable in vector form, which is DBHa (all the DBH values for A. nitens)
DBHn
#We also defined a similar variable in vector form, DBHh (which is all the DBH variables for A. hirsutus)
DBHh
#we can run a t-test just by defining one variable as x, the other as y. 
ttest_result1 = t.test(x=DBHh, y=DBHn, paired=FALSE, alternative="two.sided")
ttest_result1
```


## Grouping variables: combine two variables into one, then splitting one into two

Another way to go about doing this would be to combine the two 

```{r}
#During Digital Lab 1, we showed you how to combine two dataframes downloaded from BIEN:
#That involved the package DPLYR
library(dplyr)
#And the command rbind to create one dataframe with all the original columns for both species
Artocarpus2<-rbind(Anitidus_DBH_small, Ahirsutus_DBH_small)
head(Artocarpus2)
tail(Artocarpus2)
class(Artocarpus2$trait_value)
#Then we defined a numeric variable, combining all the DBH values for both species
DBHboth = as.numeric(Artocarpus2$trait_value)
class(DBHboth)
```

```{r}
#Plotting that without the grouping variable results in a dumb boxplot (and an always-useful histogram)
hist(DBHboth)
boxplot(DBHboth)
```

```{r}
#But We can do something clever. 
#First, combine the numeric variable DBHboth, adding it as a sixth column into the
#BIEN dataframe just created above:
ArtocarpusBoth<-cbind(Artocarpus2, DBHboth)
head(ArtocarpusBoth)
tail(ArtocarpusBoth)
```



```{r}
#Now, using the variable "scrubbed_species_binomial" divides the overall data into two groups, by species
#in the revised code below, notice the ~ symbol (known as a tilde)
#It tells R to use what comes after as a grouping variable
boxplot(DBHboth~scrubbed_species_binomial, data=ArtocarpusBoth)
```



```{r}
#similarly, we can use that same grouping trick, calling the same t-test function
#above we had used x=DBHn and x=DBHh
#but now we will use our different format - DBHboth~scrubbed_species_binomial
#again, that ~  tells R to use what comes after it as a grouping variable
ttest_result2 = t.test(DBHboth~scrubbed_species_binomial, ArtocarpusBoth, paired=F)
ttest_result2
```

####To complete this portfolio:

Before our next digital lab on Feb 22, choose one of the exercises that involve choosing two species, downloading trait data for both, and then creating boxplots and conducting t-tests.  

Also use this last coding exercise to obtain species lists for three counties in a state in the USA outside of New York. For example, you could start with the county where you attended high school. Or, you could use a county where you have vacationed in the past, or where you would like to visit. Also download data for two adjacent counties. 
