---
title: "Species lists with abundance data, and biodiversity metrics"
output: html_notebook
---

## Potentially useful: http://www.flutterbys.com.au/stats/tut/tut13.2.html

```{r}
library(vegan)
library(labdsv)
library(plyr)
```

```{r}
one=c("A","A","A","A","A","A","B","B","B","B","B","B","C","C","C","C","C","C","D","E","F","G","H")
one
length(one)
#All of this initial work is based on the example from class on Feb 13. We are creating a vector of character variables. We want to look at it, and also know the length of the vector. In this case, length gives us the number of individuals in our samples (or any species).
```



```{r}
one
one.ab=count(one)
#Now we want to figure out something about individuals within species, finding out how many individuals there are of each species. This is a measure of the abundance of each spcies. For this, we use the function "count" and we create a new variable which has the count of each species
one.ab
class(one.ab)
length(one.ab)
#We might also want to get a list of species, and the length of that list (which is species richness). 
one.u=unique(one)
one.u
length(one.u)
attributes(one.ab)

#one.pa=count(one.u)
#one.pa
```

```{R}
#We can combine two of these vectors to  compare them, which will use the R packages VEGAN and LABDSV.
#But before we do the cmobining, we must add a variable to allow keeping track of which sample is which
one.ab=data.frame("one", one.ab)
one.ab
#This added a first column with every row populated by the string variable "one"
#Let's also re-name columns. Later, our analyses of species richenss will need specifically named variables. 
names(one.ab)<-c("ListID", "SpeciesID", "Freq")
one.ab
```

three.ab=data.frame("three", three.ab)
three.ab
names(three.ab)<-c("ListID", "SpeciesID", "Freq")


```{r}
three=c("A","A","A","A","A","A","B","B","B","B","B","B","B","C","C","C","C","C","C","C","D","E","F","G","Q")
three.ab=count(three)
three.ab
#We might also want to get a list of species, and the length of that list (which is species richness). 
three.u=unique(three)
three.u
length(three.u)
```


```{R}
#We can combine two files and compare them, which is going to start using the package VEGAN and LABDSV.
#But before we do that, we need to add a variable to allow keeping track of which sample is which
#one.ab=data.frame("one", one.ab)
#one.ab
#names(one.ab)<-c("ListID", "SpeciesID", "Freq")
three.ab=data.frame("three", three.ab)
three.ab
names(three.ab)<-c("ListID", "SpeciesID", "Freq")
ab=rbind(one.ab, three.ab)
ab
dim(ab)
#A powerful function in LABDSV converts this list of sample IDs, species and site-specific abundances into a matrix in which species names are column headings, and row names are samples. To do this, it creates a master list of all species in all samples. A matrix like this is useful for quantifying multiple aspects of biodiversity -- alpha and beta and gamma. 
abundance<-matrify(ab)
abundance
#If we look at the dimensions of this matrix, what does it mean? What is 2? What is 9?
dim(abundance)
```


## After you have a matrix: diversity indices, ordinations and ordination plotting

Shannon-Weaver Diversity Index (H), possibly familiar from ecology or conservation biology classes, examines each species for the proportion of individuals that it contributes to the total individuals of all species in the sample (Pi). 

If S is the total number of species in the transect (i.e., the species richness), 
Diversity,   H  =  - SUM OVER species 1 to species i(Pi)*(ln(Pi))
							(NOTE: "ln" is the natural logarithm and i refers to the ith species)


As an example, let's return to the data above, where N total = 23 individuals:

```{r}
one
length(one)
```


where N total = 23 individuals

Let Ni be the number of individuals of any given species, and Pi = Ni/Ntotal:

Species	Ni	Pi            	ln(Pi)  	Pi * ln(Pi)
A	      6	  6/23 = 0.26   	-1.34   	-0.35
B	      6	  6/23 = 0.26   	-1.34   	-0.35
C	      6	  6/23 = 0.26   	-1.34   	-0.35
D       1   1/23 = 0.045	  -3.13    	-0.136
E       1   1/23 = 0.045	  -3.13    	-0.136
F       1   1/23 = 0.045	  -3.13    	-0.136
G       1   1/23 = 0.045	  -3.13    	-0.136
H       1   1/23 = 0.045	  -3.13    	-0.136
Q       0   0/23 = 0          0          0
				H = -(-1.73)
				
You can try doing a similar calculation for Sample "three" and then compare to the values obtained from R using the VEGAN package

```{r}
abundance
shannon<-diversity(abundance, index = "shannon")
shannon
```
				

Another measure of diversity, Simpson's index. There are actually a lot of different versions of Simpsons index, all roughly based on the likelihood that any two randomly encountered individuals are the same species. 

In VEGAN, it is calculated as

D = 1 - Sum(p^2)

where p = n/N

where each n refers to the number of organisms of that particular species
and N refers to the number of organisms of all species

Species	ni	p=n/N           p^2  
A	      6	  6/23=0.26       0.068
B	      6	  6/23=0.26       0.068
C	      6	  6/23=0.26       0.068
D	      1	  1/23=0.043      0.00189
E	      1	  1/23=0.043      0.00189
F	      1	  1/23=0.043      0.00189
G	      1	  1/23=0.043      0.00189
H	      1	  1/23=0.043      0.00189
Q       0	  0

                SUM         0.214

D = 1 - Sum(p^2) = 0.786

```{r}
abundance
simpson<-diversity(abundance, index = "simpson")
simpson
```




```{r}
#We need to add more samples if we want to move toward making multiple pairwise comparisons
two=c("A","A","A","A","A","A","B","B","B","B","B","B","B","B","C","C","C","C","C","C","D","D","D","D","D","D","D","E","F","E","E","E","E","E","F","F","F","F","F","F","F","F","F","G","G","G","G","G","G","H","H","H","H","H","H","H","H")
five=c("A","A","A","A","A","A","B","B","B","B","B","B","B","D","D","D","D","D","D","D","J","K","L","M","N")
#Here are two additional samples named "two" and "five"
#By adding to this chunk of code (and running it), see if you can add to the output above to obtain similar data
two.ab=count(two)
two.ab
#We might also want to get a list of species, and the length of that list (which is species richness). 
two.u=unique(two)
two.u
length(two.u)
five.ab=count(five)
five.ab
#We might also want to get a list of species, and the length of that list (which is species richness). 
five.u=unique(five)
five.u
length(five.u)
```


```{R}
#Let's add more data so we can continue to delve into what can be donetwo.ab
#Repeating what we did earlier, to process these lists so that we can incorporate them into an abundance matrix
two.ab=data.frame("two", two.ab)
two.ab
five.ab=data.frame("five", five.ab)
five.ab
names(two.ab)<-c("ListID", "SpeciesID", "Freq")
names(five.ab)<-c("ListID", "SpeciesID", "Freq")
one.ab
three.ab
two.ab
five.ab
ab=rbind(one.ab, three.ab, two.ab, five.ab)
ab
dim(ab)
#ab<-ab[,2:4]
abundance<-matrify(ab)
abundance
```



What can we do with this bigger matrix? For starters, we can use an index to measure the similarity (or dissimilarity) of any two pairs of lists in the matrix

There are many different ways to measure community similarity or dissimilarity. 

An enduringly useful pairwise index that uses abundance data is called the Bray-Curtis coefficient, which is a ratio. 

First, to get the top/numerator of the ratio: for all species across all samples, compute the difference in abundance between a pair of samples (as an absolute value). Then, all of those absolute differences are summed.

Next, to get the bottom/denominator of the ratio: sum the total individuals in the pair of samples

For our simple examples of samples "one" and "five": 

            "one"           "five"      |abs diff|
A 	           6	            6               0
B	             6	            7	              1 
C              6              0               6
D	             1              7	              6
E		           1              0               1
F		           1              0               1 
G		           1              0               1 
H		           1              0               1 
Q		           0              0               0 
J		           0              1               1
K		           0              1               1
L		           0              1               1
M		           0              1               1
N		           0              1               1

sum           23             25              22


Bray-Curtis = 22/(23+25) = 0.458


A simple pairwise index that uses only presence-absence data is called Jaccard's Coefficient of similarity, J. 

J = a / (n - d)

where a is the number of "positive matches" (+ +, species found in two lists), d is the number of negative matches (- - , species found in neither list) and n is the total number of possible matches (based on the total number of taxa found across all lists). 

For our simple examples of samples "one" and "five": 

            "one"           "five" 
A 	           x	            x
B	             x	            x	 
C              x
D	             x              x	
E		           x
F	             x  	
G	             x	
H	             x
Q	           	
J		                          x
K                             x
L                             x
M                             x    
N                             x


in which a = 3, d = 1, n = 14

so J  =  3 / (14 - 1) = 0.769 

This index will vary between zero and one. 

There are many other ways to measure pair-wise similarities. 

If there are more than two lists, it's possible to put together a matrix of pair-wise values for the index, which is what the code below does. 

```{r}
#the function vegdist is used on the matrix, the metric specified. The assumption is that the data are abundance data, even though some indices require binary data. So the unstated and default argument that is always used in vegdist is that  binary=FALSE. This can and should be changed if computing a presence-absence metric like Jaccard.
bray = vegdist(abundance, "bray") 
bray
jaccard = vegdist(abundance, "jaccard", binary=T)
jaccard
```



```{r}
#Now that there are four samples, it makes sense to start seeing if this is enough samples to gauge diversity across these samples. Vegan can compute the cumulative number of species across all four, and then plot them. It will take the plots in random order. Other options for this method argument include "collector" which is the order in which they are entered into the matrix. Another is "rarefaction" which will adjust to take into account that each sample may have a different number of individuals.
accurve<-specaccum(abundance, method="random", permutations=100)
plot(accurve$sites, accurve$richness,
     xlab="Number of Sites",
     ylab="Species Richness",
     main="Now with four samples")
accurve2<-specaccum(abundance, method="rarefaction", permutations=100)
plot(accurve2$sites, accurve$richness,
     xlab="Number of Sites",
     ylab="Species Richness",
     main="Rarefaction, with four samples")
```


```{r}
head(abundance)
example_NMDS=metaMDS(abundance, k=3) # Our community-by-species matrix 
# K=2 is setting the number of reduced dimensions. Increase if high stress is problem. 
#"The stress, or the disagreement between 2-D configuration and predicted values from the regression"

#A good rule of thumb: stress > 0.05 provides an excellent representation in reduced dimensions, > 0.1 is great, >0.2 is good/ok, and stress > 0.3 provides a poor representation

plot(example_NMDS)
test.plot<-ordiplot(example_NMDS,type="n") #Ordination plot function especially for congested plots
points(test.plot, "species", pch=21, col="red", bg="yellow")

test.plot<-ordiplot(example_NMDS,type="n") #Ordination plot function especially for congested plots
points(test.plot, "sites", col="blue", cex=0.9)

test.plot<-ordiplot(example_NMDS,type="n") #Ordination plot function especially for congested plots
points(test.plot, "species", pch=21, col="red", bg="yellow")
text(test.plot, "sites", col="blue", cex=0.9)

test.plot<-ordiplot(example_NMDS,type="n") #Ordination plot function especially for congested plots
text(test.plot, "species", pch=21, col="red", bg="yellow")
points(test.plot, "sites", col="blue", cex=0.9)


```





```{r}
#library("bien")
#As you know, you can obtain species occurrence data from the BIEN database, or from another even more comprehensive data base called gbif (gbif.org). Occurrence data is a type of abundance data, and it can also be used to figure out presence-absence data. 

bronx <- BIEN_occurrence_county(country="United States", state="New York", county = "Bronx")
head(bronx)
tail(bronx)
manhattan <- BIEN_occurrence_county(country="United States", state="New York", county = "New York")
head(manhattan)
tail(manhattan)
attributes(manhattan)
westchester <- BIEN_occurrence_county(country="United States", state="New York", county = "Westchester")
head(westchester)
tail(westchester)
attributes(westchester)
```
CHALLENGE THEM TO DO THEIR COUNTY LISTS

```{r}
#We previously learned to "slim down" data sets to one or a few variables, which we can do here
bronx.occ=bronx[c(1)]
head(bronx.occ)
manhattan.occ=manhattan[c(1)]
head(manhattan.occ)
westchester.occ=westchester[c(1)]
head(westchester.occ)
#These are now two species occurrence lists, one for The Bronx, one for Manhattan
#To turn them into species lists with abundance data, you can recycle the code from above.
bronx.ab=count(bronx.occ)
head(bronx.ab)
manhattan.ab=count(manhattan.occ)
head(manhattan.ab)
westchester.ab=count(westchester.occ)
head(westchester.ab)
#Then re-name the columns as previously. Here just the Bronx
bronx.ab=data.frame("bronx", bronx.ab)
head(bronx.ab)
names(bronx.ab)<-c("ListID", "SpeciesID", "Freq")
head(bronx.ab)
manhattan.ab=data.frame("manhattan", manhattan.ab)
head(manhattan.ab)
names(manhattan.ab)<-c("ListID", "SpeciesID", "Freq")
head(manhattan.ab)
westchester.ab=data.frame("westchester", westchester.ab)
head(westchester.ab)
names(westchester.ab)<-c("ListID", "SpeciesID", "Freq")
head(westchester.ab)
threecounty.ab=rbind(bronx.ab, manhattan.ab, westchester.ab)
head(threecounty.ab)
tail(threecounty.ab)
dim(threecounty.ab)
#This powerful function matrify is going to convert this list of species with their site-specific abundances into a matrix with the species as the column headings, the rows as the samples. This gives a long sample list
urban.ab<-matrify(threecounty.ab)
urban.ab
#If we look at the dimensions of this matrix, what does it mean? What is 2? What is 9?
dim(urban.ab)
simpson<-diversity(urban.ab, index = "simpson")
shannon<-diversity(urban.ab, index = "shannon")
simpson
shannon
bray = vegdist(urban.ab, "bray")
bray
jaccard = vegdist(urban.ab, "jaccard", binary=T)
jaccard


```


```{r}
#calculating a pairwise similarity or dissimilarity matrix is not that informative, if we want to move toward making multiple pairwise comparisons
two=c("A","A","A","A","A","A","B","B","B","B","B","B","B","B","C","C","C","C","C","C","D","D","D","D","D","D","D","E","F","E","E","E","E","E","F","F","F","F","F","F","F","F","F","G","G","G","G","G","G","H","H","H","H","H","H","H","H")
five=c("A","A","A","A","A","A","B","B","B","B","B","B","B","D","D","D","D","D","D","D","J","K","L","M","N")
#Here are two additional samples named "two" and "three"
#By adding to this chunk of code (and running it), see if you can add to the output above to obtain similar data
two.ab=count(two)
two.ab
#We might also want to get a list of species, and the length of that list (which is species richness). 
two.u=unique(two)
two.u
length(two.u)
five.ab=count(five)
five.ab
#We might also want to get a list of species, and the length of that list (which is species richness). 
five.u=unique(five)
five.u
length(five.u)
```





```(r)
add dune examples
```

```{r}
#do example from Manahatta
MH=read.csv("WelikiaMH.csv")
head(MH)
tail(MH)
speciesID=paste(MH$Scientific, MH$Name, sep=" ")
speciesID
MH_list=count(speciesID)
head(MH_list)
MH_list=data.frame(MH_list)
MH_list=data.frame("MH", MH_list)
names(MH_list)<-c("ListID", "SpeciesID", "Freq")
LES=read.csv("Welikiales.csv")
head(LES)
tail(LES)
speciesID=paste(LES$Scientific, LES$Name, sep=" ")
speciesID
LES_list=count(speciesID)
head(LES_list)
LES_list=data.frame("LES", LES_list)
names(LES_list)<-c("ListID", "SpeciesID", "Freq")
Welikia2=rbind(LES_list, MH_list)
dim(Welikia2)
wab=matrify(Welikia2)
wab

simpson<-diversity(wab, index = "simpson")
shannon<-diversity(wab, index = "shannon")
simpson
shannon
jaccard = vegdist(wab, "jaccard")
jaccard

```




